Formal Verification Framework Based on Symbolic Execution for Smart Contract

### 2023-12-18 papers/static-checker/ecoop2024/ecoop2024.tex
write the property check

\begin{lstlisting}[float,captionpos=b,caption={Smart contract property specification},label={lst:smart-contract-property-specification},numbers=left]
mcontract Name = spec
code := {c}
input := t
output := t'
pre-condition := cd
post-condition := cd'
\end{lstlisting}
Listing \ref{lst:smart-contract-property-specification} shows the syntax of our property specification language. Here, \KMCONTRACT, \KWITH, \KCODE, \KINPUT, \KOUTPUT, \KPRECONDI, and \KPOSTCONDI\ are keywords. The \KNAME\ is replaced by the contract name. In the input and output fields, the input and output patterns are specified. Since Michelson operates on a stack, we assume that the stack has only one element, as specified in the input pattern. Similarly, the output stack is assumed to be a stack with only one element, which is specified in the output field. As we have mentioned before, the output should be a pair consisting of an operation list and a storage. In our syntax, for verification purpose, we omit the operation list and then the output is represented by the patern of the value in the storage after the execution is finished. The input and output patterns represent Michelson terms, which can be either concrete values or symbolic values, including variables. When a variable appears for the first time, it needs to be accompanied by its corresponding type declaration. However, subsequent mentions can omit the type declaration. Formally, the pre and post-condition instances are formulas in first-order logic, and they can be handled by an SMT solver, namely the Z3 solver in our framework. The detail of the language grammar is described in in Figure \ref{}. Formulas that are used in assertions or added to solvers are terms of Boolean sort. Otherwise, terms of Boolean and non-Boolean sort may be mixed in any combination where sorts match up. Universal and existential quantifiers bind variables to the scope of the quantified formula.
\begin{lstlisting}[float,captionpos=b,caption={The specification of Add contracr},label={lst:add-contract-specification},numbers=left]
mcontract Add = spec
code := {UNPAIR; COMPARE; GE; 
        IF {ADD}{PUSH string 'Unexpected Pair'; FAILWITH}}
input := Pair (x : int) (y : int)
output := (z : int)
pre-condition := (x >= 5) | (y >= 3)
post-condition := (z >= 8)
\end{lstlisting}
Let's consider the  example of the property specification for the contract named Add as in Listing \ref{lst:add-contract-specification}. Here, the program takes as input a stack with only one element, which is a pair of two integers, symbolically specified in the Input field. It then unpairs the input pair to obtain two elements as two integers \lstinline/x/ and \lstinline/y/.  It then proceeds to destructure the input pair, yielding two integers. Subsequently, the program performs a comparison between these two integers. If  \lstinline/x/ is greater than or equal to \lstinline/y/, the program constructs a new stack with a solitary element at the top representing the sum of  \lstinline/x/ and \lstinline/y/, designated as \lstinline/z/. Otherwise, the execution terminates with an error message containing the content 'Unexpected pair.' The specification continues by defining pre and post conditions, where the precondition states the assumption that the integer  \lstinline/x/ is greater than or equal to 5 and \lstinline/y/ is greater than or equal to 3. Here, the symbol  \lstinline/|/ is the conjuration. The postcondition asserts that the result sum \lstinline/z/ is greater than or equal to 8.

@@

\begin{lstlisting}[float,captionpos=b,caption={Auction contract specification},label={lst:auction-contract-specification},numbers=left]
mcontract Auction = spec 
  entrypoint %bid
    code := {...}
    input := Pair Unit (Pair (auction_open: bool) 
                             (Pair (highest_bidder: mutez)
                                   (contract_owner: address)))
    output :=   Pair (post_auction_open: bool) 
                     Pair (post_highest_bidder: address)
                          (post_contract_owner: address))
    pre-condition := auction_open |
             Amount > Balance
    post-condition := post_highest_bidder = Sender |
             Post_Balance = Amount |
             Transfer_token (Unit, Balance, highest_bidder) |
             post_contract_owner = contract_owner ;
  entrypoint %close
    code := {...}
    input := Pair Unit (Pair (auction_open: bool) 
                             (Pair (highest_bidder: mutez)
                                   (contract_owner: address)))
    output :=   Pair (post_auction_open: bool) 
                     Pair (post_highest_bidder: address)
                          (post_contract_owner: address))
    pre-condition := auction_open |
           Sender = contract_owner                
    post-condition := not post_auction_open |
          Post_Balance = 0 |
          Transfer_token (Unit, Balance, contract_owner) ;
    (%create -> %bid) with (auction_open = true)  
                      && (Amount > Balance) | 
    (%create  -> %close) with (auction_open = true) 
                      && (Sender = contract_owner) |
    (%bid -> %bid) with (Amount > Balance)  | 
    (%bid -> %close) with (Sender = contract_owner)  | 
    not (%close -> %bid) |
    not (%close -> %close)
\end{lstlisting}

@@

\subsection{User specification}
In the user specification phase, users compose the specification in our domain-specific language using the syntax outlined in Section \ref{sec:domain-specific-language}. Note that Michelson code, whether authored by users, obtained from the Tezos blockchain, or compiled from higher-level languages, can be seamlessly employed within our framework. The specification is subsequently parsed into an Abstract Syntax Tree (AST) in OCaml. The code and the input pattern are then utilized as inputs for the interpreter during the intermediate phase to facilitate symbolic execution. In the next phase, the pre and post predicates are transformed into Z3 formulas by the Z3 converter tool.
\subsection{Michelson interpreter and Z3 converter}
The interpreter implements the execution model described in \ref{sec:symbolic-execution-model}. When provided with a concrete input, the interpreter produces a concrete value. In the presence of symbolic input, the result comprises a set of symbolic states represented as terms in OCaml. Due to Michelson being built in OCaml, the conversion of Michelson types or terms to their OCaml counterparts is straightforward. Each state contains the output stack, which has only one element consisting of an operation list and storage, along with the path conditions represented as predicates in conjunctive form. This output is then fed into the Z3 converter, which handles two different types of terms: assertion terms from the AST representing the pre and post-condition specifications, and terms resulting from the interpreter. We utilize the Z3 library in OCaml to construct this converter. Some adjustments are necessary to convert these terms into Z3 formulas. For example, as there is no natural number type (\TNAT) in Z3, the term is converted as an integer type, accompanied by constraints asserting that it is greater than or equal to zero. Several constants, such as \CAMOUNT, \CBALANCE, \CSENDER\ and cryptographic functions, such as \FSHA, \FHASHKEY\ are predefined in the environment that works for any smart contract specifications.
\subsection{Static checker}
The static checker now performs the verification process by employing symbolic execution against the property specification in the DSL. It utilizes the results from the previous phases, including symbolic execution, Z3 formulas representing pre and post conditions, and path conditions.

@@

The static checker provides a feature to generate all fail conditions implemented in a smart contract code. The tool searches for all failwith states, resulting from an execution terminated by the \FAILWITH\ instruction, from the final state space, collects the path conditions and the recorded message. The result returned to users includes the fail condition together with the message. For example, for the specification in Listing \ref{lst:add-contract-specification}, the fail condition is recorded as \lstinline/(x < y): 'Unexpected Pair'/. For the auction contract, there are several fail conditions. For the entrypoint \lstinline/bid/, the fail condition is as shown in Listing \ref{lst:fail-condition-auction-contract-bid} and for the \lstinline/close/ entrypoint, it is recorded in Listing \ref{}.

\begin{lstlisting}[float,captionpos=b,caption={Auction contract specification},label={lst:fail-condition-auction-contract},numbers=left]
1. not (auction_open): 'Already closed'
2. (auction_open = true) &&
(Amount <= Balance): 'Not higher than the current highest bid'
3. (auction_open = true) &&
(Amount <= Balance) &&
get_contract (Unit, highest_bidder) = None: 'Invalid address' 
\end{lstlisting}

\subsubsection{Property check}

consider the code is as a black box
\section {Case Studies}
\subsection{USDtz}
\subsection{Kolibri Oracle Contract}
\section {Related Work}
\section {Conclusion}
\section {Appendix}
\end{document}

>>>>

\begin{lstlisting}[float,captionpos=b,caption={Smart contract property specification},label={lst:smart-contract-property-specification},numbers=left]
mcontract Name = spec
code := {c}
input := t
output := t'
pre-condition := cd
post-condition := cd'
\end{lstlisting}
Listing \ref{lst:smart-contract-property-specification} shows the syntax of our property specification language. Here, \KMCONTRACT, \KWITH, \KCODE, \KINPUT, \KOUTPUT, \KPRECONDI, and \KPOSTCONDI\ are keywords. The \KNAME\ is replaced by the contract name. In the input and output fields, the input and output patterns are specified. Since Michelson operates on a stack, we assume that the stack has only one element, as specified in the input pattern. Similarly, the output stack is assumed to be a stack with only one element, which is specified in the output field. As we have mentioned before, the output should be a pair consisting of an operation list and a storage. In our syntax, for verification purpose, we omit the operation list and then the output is represented by the patern of the value in the storage after the execution is finished. The input and output patterns represent Michelson terms, which can be either concrete values or symbolic values, including variables. When a variable appears for the first time, it needs to be accompanied by its corresponding type declaration. However, subsequent mentions can omit the type declaration. Formally, the pre and post-condition instances are formulas in first-order logic, and they can be handled by an SMT solver, namely the Z3 solver in our framework. The detail of the language grammar is described in in Figure \ref{}. Formulas that are used in assertions or added to solvers are terms of Boolean sort. Otherwise, terms of Boolean and non-Boolean sort may be mixed in any combination where sorts match up. Universal and existential quantifiers bind variables to the scope of the quantified formula.
\begin{lstlisting}[float,captionpos=b,caption={The specification of Add contracr},label={lst:add-contract-specification},numbers=left]
mcontract Add = spec
code := {UNPAIR; COMPARE; GE; 
        IF {ADD}{PUSH string 'Unexpected Pair'; FAILWITH}}
input := Pair (x : int) (y : int)
output := (z : int)
pre-condition := (x >= 5) | (y > 3)
post-condition := (z >= 8)
\end{lstlisting}
Let's consider the  example of the property specification for the contract named Add as in Listing \ref{lst:add-contract-specification}. Here, the program takes as input a stack with only one element, which is a pair of two integers, symbolically specified in the Input field. It then unpairs the input pair to obtain two elements as two integers \lstinline/x/ and \lstinline/y/.  It then proceeds to destructure the input pair, yielding two integers. Subsequently, the program performs a comparison between these two integers. If  \lstinline/x/ is greater than or equal to \lstinline/y/, the program constructs a new stack with a solitary element at the top representing the sum of  \lstinline/x/ and \lstinline/y/, designated as \lstinline/z/. Otherwise, the execution terminates with an error message containing the content 'Unexpected pair.' The specification continues by defining pre and post conditions, where the precondition states the assumption that the integer  \lstinline/x/ is greater than or equal to 5 and \lstinline/y/ is greater than or equal to 3. Here, the symbol  \lstinline/|/ is the conjuration. The postcondition asserts that the result sum \lstinline/z/ is greater than or equal to 8.

@@

\begin{lstlisting}[float,captionpos=b,caption={Auction contract specification},label={lst:auction-contract-specification},numbers=left]
mcontract Auction = spec 
  entrypoint %bid
    code := {...}
    input := Pair Unit (Pair (auction_open: bool) 
                             (Pair (highest_bidder: address)
                                   (contract_owner: address)))
    output :=   Pair (post_auction_open: bool) 
                     Pair (post_highest_bidder: address)
                          (post_contract_owner: address))
    pre-condition := auction_open |
             Amount > Balance
    post-condition := post_highest_bidder = Sender |
             Post_Balance = Amount |
             Transfer_token (Unit, Balance, highest_bidder) |
             post_contract_owner = contract_owner ;
  entrypoint %close
    code := {...}
    input := Pair Unit (Pair (auction_open: bool) 
                             (Pair (highest_bidder: address)
                                   (contract_owner: address)))
    output :=   Pair (post_auction_open: bool) 
                     Pair (post_highest_bidder: address)
                          (post_contract_owner: address))
    pre-condition := auction_open |
           Sender = contract_owner                
    post-condition := not post_auction_open |
          Post_Balance = 0 |
          Transfer_token (Unit, Balance, contract_owner) ;
    (%create -> %bid) with (auction_open = true)  
                      && (Amount > Balance) | 
    (%create  -> %close) with (auction_open = true) 
                      && (Sender = contract_owner) |
    (%bid -> %bid) with (Amount > Balance)  | 
    (%bid -> %close) with (Sender = contract_owner)  | 
    not (%close -> %bid) |
    not (%close -> %close)
\end{lstlisting}

@@

\subsection{User specification}
In the user specification phase, users compose the specification in our domain-specific language using the syntax outlined in Section \ref{sec:domain-specific-language}. Note that Michelson code, whether authored by users, obtained from the Tezos blockchain, or compiled from higher-level languages, can be seamlessly employed within our framework. The specification is subsequently parsed into an Abstract Syntax Tree (AST) in OCaml. The code and the input pattern are then utilized as inputs for the interpreter during the intermediate phase to facilitate symbolic execution. In the next phase, the pre and post predicates are transformed into Z3 formulas by the Z3 converter tool.
\subsection{Michelson interpreter and Z3 converter}
The interpreter implements the execution model described in \ref{sec:symbolic-execution-model}. When provided with a concrete input, the interpreter produces a concrete value. In the presence of symbolic input, the result comprises a set of symbolic states represented as terms in OCaml. Due to Michelson being built in OCaml, the conversion of Michelson types or terms to their OCaml counterparts is straightforward. Each state contains the output stack, which has only one element consisting of an operation list and storage, along with the path conditions represented as predicates in conjunctive form. This output is then fed into the Z3 converter, which handles two different types of terms: assertion terms from the AST representing the pre and post-condition specifications, and terms resulting from the interpreter. We utilize the Z3 library in OCaml to construct this converter. Some adjustments are necessary to convert these terms into Z3 formulas. For example, as there is no natural number type (\TNAT) in Z3, the term is converted as an integer type, accompanied by constraints asserting that it is greater than or equal to zero. Several constants, such as \CAMOUNT, \CBALANCE, \CSENDER\ and cryptographic functions, such as \FSHA, \FHASHKEY\ are predefined in the environment that works for any smart contract specifications.

\paragraph{Dealing with loop formalization}
As mentioned earlier, our execution model formalizes loops symbolically using the \FOLD\ and \FMAP\ functions. This approach has the advantage of allowing us to convert these loops into formulas compatible with certain SMT solvers, such as Z3, which support these functions. Specifically, Z3 provides \lstinline/seq.map fn s/, a map function on sequence \lstinline/s/, and \lstinline/(seq.foldl fn b s)/, a fold function with an initial value \lstinline/b/ on sequence \lstinline/s/. These functions are supported for checking assertions that contain them. Therefore, we can directly convert the symbolic results of loops into formulas in Z3.
\subsection{Static checker}
The static checker now performs the verification process by employing symbolic execution against the property specification in the DSL. It utilizes the results from the previous phases, including symbolic execution, Z3 formulas representing pre and post conditions, and path conditions.

@@

The static checker provides a feature to generate all fail conditions implemented in a smart contract code. The tool searches for all failwith states, resulting from an execution terminated by the \FAILWITH\ instruction, from the final state space, collects the path conditions and the recorded message. The result returned to users includes the fail condition together with the message. For example, for the specification in Listing \ref{lst:add-contract-specification}, the fail condition is recorded as \lstinline/(x < y): 'Unexpected Pair'/.

For the auction contract, there are several fail conditions. For the entrypoint \lstinline/bid/, the fail condition is as shown in Listing \ref{lst:fail-condition-auction-contract-bid} and for the \lstinline/close/ entrypoint, it is recorded in Listing \ref{lst:fail-condition-auction-contract-close}. For the \lstinline/bid/ entrypoint, there are three fail conditions. First is if the auction is closed, and then if it is not, the contract checks if the amount is greater than the balance. If not, the call fails as well. The last fail happens when the auction is open, the amount is greater than the balance, but the highest bidder's address is invalid. Note that there are several addresses that are valid as Michelson addresses but are not associated with any real account on the Tezos blockchain, or the parameter's type is not correct, while the transfer can only succeed if it is the given parameter and the address is associated with an account with the correct parameter on the Tezos blockchain. Bidding returns the previous highest bid as part of the transaction where the highest bidder is stored as a value of type address. However, to receive a token transfer, this address must be cast (by the Michelson implementation of the contract) to an implicit contract of Michelson type (\TCONTRACT\ \TUNIT). This cast is unavoidable, as a value of contract type cannot be stored. However, the cast may fail, and its failure leads to an error condition that is reported with a \FAILWITH\ instruction in this implementation. In our framework, this cast fail case is shown in the fail condition, but in the user preconditions, it assumes that they are successfully casted. The \lstinline/close/ entrypoint has similar fail conditions when the auction is closed or the owner's address is uncastable. But the other fail condition is when the auction is open, but the sender who calls this entrypoint is not the owner of the contract, which was stored in the contract owner part.

\begin{lstlisting}[float,captionpos=b,caption={Fail condition for Auction contract bid entrypoint },label={lst:fail-condition-auction-contract-bid},numbers=left]
1. not (auction_open): 'Already closed'
2. (auction_open = true) &&
(Amount <= Balance): 'Not higher than the current highest bid'
3. (auction_open = true) &&
(Amount <= Balance) &&
get_contract (Unit, highest_bidder) = None: 'Invalid address' 
\end{lstlisting}

\begin{lstlisting}[float,captionpos=b,caption={Fail condition for Auction contract close entrypoint},label={lst:fail-condition-auction-contract-close},numbers=left]
1. Not (auction_open): 'Already closed'
2. (auction_open = true) &&
Sender != contract_owner): 'Not owner'
3.  (auction_open = true) &&
(Sender = contract_owner) &&
get_contract (Unit, contract_owner) = None: 'Invalid address'
\end{lstlisting}

In this implementation of the auction contract, the failures are recorded as expected. The contract is designed to fail when the auction is closed, the bid is not greater than the old highest bid, or the caller is not the contract owner when attempting to close the contract as transferring the contract balance to their account as a part of the action.

\subsubsection{Property check}
The static checker enables users to verify smart contract properties by providing pre and post condition specifications. A property is specified with the following clause:  \lstinline/pre-condtion : =/ $\mid$ $\Phi_{i}$ \lstinline/ post-condtion : =/   $\mid$ $\Psi_{j}$. This specification asserts that if all preconditions $\Phi_{i}$ hold before smart contract execution, then all $\Psi_{j}$ must become true after the execution. To verify this property, we explore all the paths $\Pi$ with the path condition $\Theta$ such that ($\bigwedge \Phi_{i}$ $\land$ $\Theta$) is satisfiable. We then check that for every path $\Pi$, $\neg$ ($\bigwedge \Phi_{i} \rightarrow \bigwedge \Psi_{j}$) is unsatisfiable in the context given by the input for the entry point and its corresponding final state and path condition.

Consider the specification of the Add contract in Listing \ref{lst:add-contract-specification}. In this context, the two variables \lstinline/x/ and \lstinline/y/ are integers. The results of the symbolic execution are either a failwith state with the path condition as the predicate stating that the variable \lstinline/x/ is less than \lstinline/y/, or a state in which the final stack has only one element, which is a symbolic variable \lstinline/sbvar/ of type integer, with the path condition stating that the variable \lstinline/x/ is greater than or equal to \lstinline/y/ and the symbolic variable \lstinline/sbvar/ is equal to the sum of \lstinline/x/ and \lstinline/y/. For this specification, the failwith branch is discarded because the precondition is not satisfy with tha path condition, and then \lstinline/z/ is matched to the symbolic variable \lstinline/sbvar/ in the other branch. In this example, the checker then verifies that given the with the context asserting that (\lstinline/x/ >= \lstinline/y/) $\land$ (\lstinline/z/ = \lstinline/x/ + \lstinline/y/), $\neg$ ((\lstinline/x/ >= 5) $\land$ (\lstinline/y/ > 3) $\rightarrow$ (\lstinline/z/ >= 8)) is unsatisfiable.

\subsubsection{Life-cycle check}
\section {Case Studies}
\subsection{USDtz}

\subsection{Kolibri Oracle Contract}
\section {Related Work}
\section {Conclusion}
\section {Appendix}
\end{document}
